---
title: "Synergus Trait Modeling Script"
author: "G. Brown"
date: "2023-02-28"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Purpose: Computationally simulate stochastic mapping of ancestral states in the Synergus phylogeny using previously established methods for deciding models.

#### Initialization

```{r message=TRUE, warning=FALSE}
library(ape)
library(phytools)
library(geiger)
library(expm)
library(lmtest)
```

#### Path to data

#### These can be adjusted based on where you are saving your files (if you are trying to recreate this analysis).

```{r warning=FALSE, include=FALSE}
setwd("//wsl.localhost/Ubuntu/home/guerbrown/github_local/biol-4386-course-project-guerbrown/data/TraitModeling")
tree <- "//wsl.localhost/Ubuntu/home/guerbrown/github_local/biol-4386-course-project-guerbrown/data/TraitModeling/inqcophy_traitmodeling_SynTree_guerbrown.nex"
traits <- "//wsl.localhost/Ubuntu/home/guerbrown/github_local/biol-4386-course-project-guerbrown/data/TraitModeling/inqcophy_traitmodeling_TraitData_guerbrown.csv"
```

### Data Import

```{r include=FALSE}
# Import
tree <- readNexus(tree)
# Rooting the phylogeny based on a version of the tree with an outgroup and plotting
root.phylo(tree, 28, resolve.root = T)
# Importing the .csv file with our traits
data <- read.csv(traits, row.names = 1)
```

### Purpose: Computationally develop models for the following traits of Synergus:
- oak section
- cynip generation
- cell type
- cluster
- gall size
- detatchable
- spines
- texture
- hair
- wool
- nectar
- rolling
- fiber
- bract
- woody 
- fleshy
- location
- season

### Why: To develop a model under which we can estimate ancestral traits for each of these characteristics. Moreover, it will provide us with a means of understanding the constraints of Synergus evolution.

### Misc.: Data run on the University of Iowa Interactive Data Analytics Service with 32 cpu cores and 128gb of RAM. 

### Modeling Procedures

**Oak Section**

```{r}
#move the data from one column into a simple matrix with the row names
oaksection <- setNames(data$oak_section, rownames(data))
#here is the previous step visualized
oaksection
#set up your models, take time to check before running. To hide the fitted model, use quiet = T. Here I have left one fit visible but will hide the remaining fits.
oaksection.ER <- fitpolyMk(tree, oaksection, model = "ER")
oaksection.SYM <- fitpolyMk(tree, oaksection, model = "SYM", quiet = T)
oaksection.ARD <- fitpolyMk(tree, oaksection, model = "ARD", quiet = T)
oaksection.transient <- fitpolyMk(tree, oaksection, model = "transient", quiet = T)
#Create a table to compare the models
data.frame(model = c("ER", "SYM", "ARD", "transient"),
           logLik = c(logLik(oaksection.ER), logLik(oaksection.SYM), 
                      logLik(oaksection.ARD), logLik(oaksection.transient)),
           k = c(attr(logLik(oaksection.ER), "df"),
                 attr(logLik(oaksection.SYM), "df"),
                 attr(logLik(oaksection.ARD), "df"),
                 attr(logLik(oaksection.transient), "df")),
           AIC = aic <- c(AIC(oaksection.ER), AIC(oaksection.SYM), 
                          AIC(oaksection.ARD), AIC(oaksection.transient)),
           weight = unclass(aic.w(aic)))
#To help anyone else interpret this, the aic value represents the 
plot(oaksection.ER)
```

**Cynip Generation**

```{r}
cynipgen <- setNames(data$cynip_generation, rownames(data))
cynipgen.ER <- fitpolyMk(tree, cynipgen, model = "ER")
cynipgen.SYM <- fitpolyMk(tree, cynipgen, model = "SYM", quiet = T)
cynipgen.ARD <- fitpolyMk(tree, cynipgen, model = "ARD", quiet = T)
cynipgen.transient <- fitpolyMk(tree, cynipgen, model = "transient", quiet = T)
data.frame(model = c("ER", "SYM", "ARD", "transient"),
           logLik = c(logLik(cynipgen.ER), logLik(cynipgen.SYM), 
                      logLik(cynipgen.ARD), logLik(cynipgen.transient)),
           k = c(attr(logLik(cynipgen.ER), "df"),
                 attr(logLik(cynipgen.SYM), "df"),
                 attr(logLik(cynipgen.ARD), "df"),
                 attr(logLik(cynipgen.transient), "df")),
           AIC = aic <- c(AIC(cynipgen.ER), AIC(cynipgen.SYM), 
                          AIC(cynipgen.ARD), AIC(cynipgen.transient)),
           weight = unclass(aic.w(aic)))
```

**mono-polythalamous**

```{r}
monopoly <- setNames(data$mono.polythalamous, rownames(data))
monopoly.ER <- fitpolyMk(tree, monopoly, model = "ER")
monopoly.SYM <- fitpolyMk(tree, monopoly, model = "SYM", quiet = T)
monopoly.ARD <- fitpolyMk(tree, monopoly, model = "ARD", quiet = T)
monopoly.transient <- fitpolyMk(tree, monopoly, model = "transient", quiet = T)
data.frame(model = c("ER", "SYM", "ARD", "transient"),
           logLik = c(logLik(monopoly.ER), logLik(monopoly.SYM), 
                      logLik(monopoly.ARD), logLik(monopoly.transient)),
           k = c(attr(logLik(monopoly.ER), "df"),
                 attr(logLik(monopoly.SYM), "df"),
                 attr(logLik(monopoly.ARD), "df"),
                 attr(logLik(monopoly.transient), "df")),
           AIC = aic <- c(AIC(monopoly.ER), AIC(monopoly.SYM), 
                          AIC(monopoly.ARD), AIC(monopoly.transient)),
           weight = unclass(aic.w(aic)))
```

### Now that I have shown 3 examples of the procedure, I will specify each chunk to just show the results. This will cut down on some space in the knitted files.

**gall Size**

```{r echo=FALSE}
gallsize <- setNames(data$gall_size, rownames(data))
gallsize.ER <- fitpolyMk(tree, gallsize, model = "ER")
gallsize.SYM <- fitpolyMk(tree, gallsize, model = "SYM", quiet = T)
gallsize.ARD <- fitpolyMk(tree, gallsize, model = "ARD", quiet = T)
gallsize.transient <- fitpolyMk(tree, gallsize, model = "transient", quiet = T)
data.frame(model = c("ER", "SYM", "ARD", "transient"),
           logLik = c(logLik(gallsize.ER), logLik(gallsize.SYM), 
                      logLik(gallsize.ARD), logLik(gallsize.transient)),
           k = c(attr(logLik(gallsize.ER), "df"),
                 attr(logLik(gallsize.SYM), "df"),
                 attr(logLik(gallsize.ARD), "df"),
                 attr(logLik(gallsize.transient), "df")),
           AIC = aic <- c(AIC(gallsize.ER), AIC(gallsize.SYM), 
                          AIC(gallsize.ARD), AIC(gallsize.transient)),
           weight = unclass(aic.w(aic)))
```

**cluster**

```{r echo=FALSE}
cluster <- setNames(data$cluster, rownames(data))
cluster.ER <- fitpolyMk(tree, cluster, model = "ER")
cluster.SYM <- fitpolyMk(tree, cluster, model = "SYM", quiet = T)
cluster.ARD <- fitpolyMk(tree, cluster, model = "ARD", quiet = T)
cluster.transient <- fitpolyMk(tree, cluster, model = "transient", quiet = T)
data.frame(model = c("ER", "SYM", "ARD", "transient"),
           logLik = c(logLik(cluster.ER), logLik(cluster.SYM), 
                      logLik(cluster.ARD), logLik(cluster.transient)),
           k = c(attr(logLik(cluster.ER), "df"),
                 attr(logLik(cluster.SYM), "df"),
                 attr(logLik(cluster.ARD), "df"),
                 attr(logLik(cluster.transient), "df")),
           AIC = aic <- c(AIC(cluster.ER), AIC(cluster.SYM), 
                          AIC(cluster.ARD), AIC(cluster.transient)),
           weight = unclass(aic.w(aic)))
```

**detachable**

```{r echo=FALSE}
detatchable <- setNames(data$detachable, rownames(data))
detatchable.ER <- fitpolyMk(tree, detatchable, model = "ER")
detatchable.SYM <- fitpolyMk(tree, detatchable, model = "SYM", quiet = T)
detatchable.ARD <- fitpolyMk(tree, detatchable, model = "ARD", quiet = T)
detatchable.transient <- fitpolyMk(tree, detatchable, model = "transient", quiet = T)
data.frame(model = c("ER", "SYM", "ARD", "transient"),
           logLik = c(logLik(detatchable.ER), logLik(detatchable.SYM), 
                      logLik(detatchable.ARD), logLik(detatchable.transient)),
           k = c(attr(logLik(detatchable.ER), "df"),
                 attr(logLik(detatchable.SYM), "df"),
                 attr(logLik(detatchable.ARD), "df"),
                 attr(logLik(detatchable.transient), "df")),
           AIC = aic <- c(AIC(detatchable.ER), AIC(detatchable.SYM), 
                          AIC(detatchable.ARD), AIC(detatchable.transient)),
           weight = unclass(aic.w(aic)))
```

**spines**

```{r echo=FALSE}
spines <- setNames(data$spines, rownames(data))
spines.ER <- fitpolyMk(tree, spines, model = "ER")
spines.SYM <- fitpolyMk(tree, spines, model = "SYM", quiet = T)
spines.ARD <- fitpolyMk(tree, spines, model = "ARD", quiet = T)
spines.transient <- fitpolyMk(tree, spines, model = "transient", quiet = T)
data.frame(model = c("ER", "SYM", "ARD", "transient"),
           logLik = c(logLik(spines.ER), logLik(spines.SYM), 
                      logLik(spines.ARD), logLik(spines.transient)),
           k = c(attr(logLik(spines.ER), "df"),
                 attr(logLik(spines.SYM), "df"),
                 attr(logLik(spines.ARD), "df"),
                 attr(logLik(spines.transient), "df")),
           AIC = aic <- c(AIC(spines.ER), AIC(spines.SYM), 
                          AIC(spines.ARD), AIC(spines.transient)),
           weight = unclass(aic.w(aic)))
```

**texture**

```{r echo=FALSE}
texture <- setNames(data$texture, rownames(data))
texture.ER <- fitpolyMk(tree, texture, model = "ER")
texture.SYM <- fitpolyMk(tree, texture, model = "SYM", quiet = T)
texture.ARD <- fitpolyMk(tree, texture, model = "ARD", quiet = T)
texture.transient <- fitpolyMk(tree, texture, model = "transient", quiet = T)
data.frame(model = c("ER", "SYM", "ARD", "transient"),
           logLik = c(logLik(texture.ER), logLik(texture.SYM), 
                      logLik(texture.ARD), logLik(texture.transient)),
           k = c(attr(logLik(texture.ER), "df"),
                 attr(logLik(texture.SYM), "df"),
                 attr(logLik(texture.ARD), "df"),
                 attr(logLik(texture.transient), "df")),
           AIC = aic <- c(AIC(texture.ER), AIC(texture.SYM), 
                          AIC(texture.ARD), AIC(texture.transient)),
           weight = unclass(aic.w(aic)))
```

**hair**

```{r echo=FALSE}
hairs <- setNames(data$hairs, rownames(data))
hairs.ER <- fitpolyMk(tree, hairs, model = "ER")
hairs.SYM <- fitpolyMk(tree, hairs, model = "SYM", quiet = T)
hairs.ARD <- fitpolyMk(tree, hairs, model = "ARD", quiet = T)
hairs.transient <- fitpolyMk(tree, hairs, model = "transient", quiet = T)
data.frame(model = c("ER", "SYM", "ARD", "transient"),
           logLik = c(logLik(hairs.ER), logLik(hairs.SYM), 
                      logLik(hairs.ARD), logLik(hairs.transient)),
           k = c(attr(logLik(hairs.ER), "df"),
                 attr(logLik(hairs.SYM), "df"),
                 attr(logLik(hairs.ARD), "df"),
                 attr(logLik(hairs.transient), "df")),
           AIC = aic <- c(AIC(hairs.ER), AIC(hairs.SYM), 
                          AIC(hairs.ARD), AIC(hairs.transient)),
           weight = unclass(aic.w(aic)))
```

**wool**

```{r echo=FALSE}
wool <- setNames(data$wool, rownames(data))
wool.ER <- fitMk(tree, wool, model = "ER")
wool.SYM <- fitMk(tree, wool, model = "SYM", quiet = T)
wool.ARD <- fitMk(tree, wool, model = "ARD", quiet = T)
data.frame(model = c("ER", "SYM", "ARD"),
           logLik = c(logLik(wool.ER), logLik(wool.SYM), 
                      logLik(wool.ARD),
           k = c(attr(logLik(wool.ER), "df"),
                 attr(logLik(wool.SYM), "df"),
                 attr(logLik(wool.ARD), "df")),
           AIC = aic <- c(AIC(wool.ER), AIC(wool.SYM), 
                          AIC(wool.ARD)),
           weight = unclass(aic.w(aic))))
```

**nectar**

```{r echo=FALSE}
nectar <- setNames(data$nectar, rownames(data))
nectar.ER <- fitpolyMk(tree, nectar, model = "ER")
nectar.SYM <- fitpolyMk(tree, nectar, model = "SYM", quiet = T)
nectar.ARD <- fitpolyMk(tree, nectar, model = "ARD", quiet = T)
nectar.transient <- fitpolyMk(tree, nectar, model = "transient", quiet = T)
data.frame(model = c("ER", "SYM", "ARD", "transient"),
           logLik = c(logLik(nectar.ER), logLik(nectar.SYM), 
                      logLik(nectar.ARD), logLik(nectar.transient)),
           k = c(attr(logLik(nectar.ER), "df"),
                 attr(logLik(nectar.SYM), "df"),
                 attr(logLik(nectar.ARD), "df"),
                 attr(logLik(nectar.transient), "df")),
           AIC = aic <- c(AIC(nectar.ER), AIC(nectar.SYM), 
                          AIC(nectar.ARD), AIC(nectar.transient)),
           weight = unclass(aic.w(aic)))
```

##### Rolling is skipped because it has only one trait shown across all taxa being studied

**fiber**

```{r echo=FALSE}
fiber <- setNames(data$fiber, rownames(data))
fiber.ER <- fitpolyMk(tree, fiber, model = "ER")
fiber.SYM <- fitpolyMk(tree, fiber, model = "SYM", quiet = T)
fiber.ARD <- fitpolyMk(tree, fiber, model = "ARD", quiet = T)
fiber.transient <- fitpolyMk(tree, fiber, model = "transient", quiet = T)
data.frame(model = c("ER", "SYM", "ARD", "transient"),
           logLik = c(logLik(fiber.ER), logLik(fiber.SYM), 
                      logLik(fiber.ARD), logLik(fiber.transient)),
           k = c(attr(logLik(fiber.ER), "df"),
                 attr(logLik(fiber.SYM), "df"),
                 attr(logLik(fiber.ARD), "df"),
                 attr(logLik(fiber.transient), "df")),
           AIC = aic <- c(AIC(fiber.ER), AIC(fiber.SYM), 
                          AIC(fiber.ARD), AIC(fiber.transient)),
           weight = unclass(aic.w(aic)))
```

**bract**

```{r echo=FALSE}
bract <- setNames(data$bract, rownames(data))
bract.ER <- fitpolyMk(tree, bract, model = "ER")
bract.SYM <- fitpolyMk(tree, bract, model = "SYM", quiet = T)
bract.ARD <- fitpolyMk(tree, bract, model = "ARD", quiet = T)
bract.transient <- fitpolyMk(tree, bract, model = "transient", quiet = T)
data.frame(model = c("ER", "SYM", "ARD", "transient"),
           logLik = c(logLik(bract.ER), logLik(bract.SYM), 
                      logLik(bract.ARD), logLik(bract.transient)),
           k = c(attr(logLik(bract.ER), "df"),
                 attr(logLik(bract.SYM), "df"),
                 attr(logLik(bract.ARD), "df"),
                 attr(logLik(bract.transient), "df")),
           AIC = aic <- c(AIC(bract.ER), AIC(bract.SYM), 
                          AIC(bract.ARD), AIC(bract.transient)),
           weight = unclass(aic.w(aic)))
```

**woody**

```{r echo=FALSE}
woody <- setNames(data$woody, rownames(data))
woody.ER <- fitpolyMk(tree, woody, model = "ER")
woody.SYM <- fitpolyMk(tree, woody, model = "SYM", quiet = T)
woody.ARD <- fitpolyMk(tree, woody, model = "ARD", quiet = T)
woody.transient <- fitpolyMk(tree, woody, model = "transient", quiet = T)
data.frame(model = c("ER", "SYM", "ARD", "transient"),
           logLik = c(logLik(woody.ER), logLik(woody.SYM), 
                      logLik(woody.ARD), logLik(woody.transient)),
           k = c(attr(logLik(woody.ER), "df"),
                 attr(logLik(woody.SYM), "df"),
                 attr(logLik(woody.ARD), "df"),
                 attr(logLik(woody.transient), "df")),
           AIC = aic <- c(AIC(woody.ER), AIC(woody.SYM), 
                          AIC(woody.ARD), AIC(woody.transient)),
           weight = unclass(aic.w(aic)))
```

**fleshy**

```{r echo=FALSE}
fleshy <- setNames(data$fleshy, rownames(data))
fleshy.ER <- fitpolyMk(tree, fleshy, model = "ER")
fleshy.SYM <- fitpolyMk(tree, fleshy, model = "SYM", quiet = T)
fleshy.ARD <- fitpolyMk(tree, fleshy, model = "ARD", quiet = T)
fleshy.transient <- fitpolyMk(tree, fleshy, model = "transient", quiet = T)
data.frame(model = c("ER", "SYM", "ARD", "transient"),
           logLik = c(logLik(fleshy.ER), logLik(fleshy.SYM), 
                      logLik(fleshy.ARD), logLik(fleshy.transient)),
           k = c(attr(logLik(fleshy.ER), "df"),
                 attr(logLik(fleshy.SYM), "df"),
                 attr(logLik(fleshy.ARD), "df"),
                 attr(logLik(fleshy.transient), "df")),
           AIC = aic <- c(AIC(fleshy.ER), AIC(fleshy.SYM), 
                          AIC(fleshy.ARD), AIC(fleshy.transient)),
           weight = unclass(aic.w(aic)))
```

**location**

```{r echo=FALSE}
location <- setNames(data$location, rownames(data))
location.ER <- fitpolyMk(tree, location, model = "ER")
location.SYM <- fitpolyMk(tree, location, model = "SYM", quiet = T)
location.ARD <- fitpolyMk(tree, location, model = "ARD", quiet = T)
location.transient <- fitpolyMk(tree, location, model = "transient", quiet = T)
data.frame(model = c("ER", "SYM", "ARD", "transient"),
           logLik = c(logLik(location.ER), logLik(location.SYM), 
                      logLik(location.ARD), logLik(location.transient)),
           k = c(attr(logLik(location.ER), "df"),
                 attr(logLik(location.SYM), "df"),
                 attr(logLik(location.ARD), "df"),
                 attr(logLik(location.transient), "df")),
           AIC = aic <- c(AIC(location.ER), AIC(location.SYM), 
                          AIC(location.ARD), AIC(location.transient)),
           weight = unclass(aic.w(aic)))
```

**season**

```{r echo=FALSE}
season <- setNames(data$season, rownames(data))
season.ER <- fitpolyMk(tree, season, model = "ER")
season.SYM <- fitpolyMk(tree, season, model = "SYM", quiet = T)
season.ARD <- fitpolyMk(tree, season, model = "ARD", quiet = T)
season.transient <- fitpolyMk(tree, season, model = "transient", quiet = T)
data.frame(model = c("ER", "SYM", "ARD", "transient"),
           logLik = c(logLik(season.ER), logLik(season.SYM), 
                      logLik(season.ARD), logLik(season.transient)),
           k = c(attr(logLik(season.ER), "df"),
                 attr(logLik(season.SYM), "df"),
                 attr(logLik(season.ARD), "df"),
                 attr(logLik(season.transient), "df")),
           AIC = aic <- c(AIC(season.ER), AIC(season.SYM), 
                          AIC(season.ARD), AIC(season.transient)),
           weight = unclass(aic.w(aic)))
```
