---
title: "Synergus Trait Modeling Script"
author: "G. Brown"
date: "2023-02-28"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Purpose: Computationally simulate stochastic mapping of ancestral states in the Synergus inquiline phylogeny using previously established methods for deciding models.

##### Initialization
```{r message=TRUE, warning=TRUE}
library(ape)
library(phytools)
library(geiger)
library(corHMM)
```

##### Path to data

```{r}
setwd("//wsl.localhost/Ubuntu/home/guerbrown/github_local/biol-4386-course-project-guerbrown/data/TraitModeling")
tree <- "//wsl.localhost/Ubuntu/home/guerbrown/github_local/biol-4386-course-project-guerbrown/data/TraitModeling/inqcophy_traitmodeling_SynTree_guerbrown.nex"
traits <- "//wsl.localhost/Ubuntu/home/guerbrown/github_local/biol-4386-course-project-guerbrown/data/TraitModeling/inqcophy_traitmodeling_TraitData_guerbrown.csv"
```

##### Synergus Tree Import

```{r}
#Import
S <- readNexus(tree)

#Rooting the phylogeny based on a version of the tree with an outgroup and plotting
root.phylo(S, 1, resolve.root = T)
plot.phylo(S, align.tip.label = T, use.edge.length = F, root.edge = T)
```

##### Data Import

```{r}
data <- read.csv(traits, row.names = 1)
```
***

### Ancestral State Reconstructions
##### To do this, I will be using the most likely model for the reconstruction of each trait. These models were decided using a seperate script, which is also provided in this folder. This script will not be run by default as it is extremely computationally expensive. I would recommend running it as a job through an HPC.  
##### One additional note, some of these models will be consider variables discrete and other polymorphic. This was decided by G. Brown and A. Forbes depending on ecology, parsimony, and state reconstruction accuracy. 

#### Oak Section 

##### Model Recapitulation

```{r}
oaksection <- setNames(data$oak_section, rownames(data))
# Rename the levels of the variable 
levels(oaksection) <- unique(data$oak_section)
# Recapitulation of the best fit model
fitER <- ace(oaksection, S, model = "ER", type = "discrete")
S$tip.label
data$oak_section
```

##### Ancestral State Reconstruction

```{r fig.height=10, fig.width=20, fig.align='center'}
svg("/home/guerbrown/github_local/biol-4386-course-project-guerbrown/output/inqcophy_traitmodeling_OakSectionRecon_guerbrown.svg", width = 20, height = 10)
cols1 <- c("red", "blue", "green", "black")
plot.phylo(S, type = "fan", cex = 2, align.tip.label = F, lwd = 3.5, label.offset = 0.1, use.edge.length = F, open.angle =  165)
sectiontotree <- setNames(data$oak_section, data$Synergus)
xx <- strsplit(as.character(sectiontotree), split = "+", fixed = T)
pp <- matrix(0, length(sectiontotree), 3, dimnames = list(names(sectiontotree), c("Quercus", "Lobatae","Virentes")))
for (i in 1:nrow(pp)) pp[i,xx[[i]]]<-1/length(xx[[i]])
par(fg = "transparent")
ape::tiplabels(pie = pp, piecol = , cex = 0.15)
par(fg = "black")
tiplabels(pie = to.matrix(oaksection, sort(unique(oaksection))),piecol = cols1, cex = 0.15)
nodelabels( node = 1:S$Nnode+Ntip(S),
    pie = fitER$lik.anc, piecol = cols1, cex=0.25)
tiplabels(pie = to.matrix(oaksection,sort(unique(oaksection))), piecol = cols1, cex = 0.15)
legend("left", c("Lobatae","Quercus", "Virentes", "Quercus+Virentes"), pch = 21, pt.bg = c("red", "blue", "green", "black"), pt.cex = 3, title = "Oak Section", cex = 1.5)
dev.off()
```

#### Cynip Generation

##### Model Recapitulation

```{r}
cynipgen <- setNames(data$cynip_generation, rownames(data))
# Rename the levels of the variable 
levels(cynipgen) <- unique(data$cynip_generation)
# Recapitulation of the best fit model
cynipgen.ER <- fitpolyMk(S, cynipgen, model = "ER")
#pull states off the fitted model object
xx2 <- apply(cynipgen.ER$data,1,
            function(x,ss) ss[which(x==1)],
            ss=colnames(cynipgen.ER$data))

#corHMM does not like to work with + so let's replace those with /
xx2<-gsub("+","/",xx2,fixed=TRUE)
#xx<-gsub("+","/",xx,fixed=TRUE)
#build corHMM data frame
gen.data <- data.frame(Genus_sp = names(xx2), cynipgen = xx2)
gen.data
#Pull out model design matrix from the fitted object
rate.mat2 <- cynipgen.ER$index.matrix
rate.mat2[rate.mat2==0] <- NA
colnames(rate.mat2) <- rownames(rate.mat2) <- gsub("+","/",
                                             colnames(rate.mat2),fixed=TRUE)
ind <- order(colnames(rate.mat2))
rate.mat <- rate.mat2[ind,ind]
cynipgen.ER$rates
#Ancestral States
fit.marginal2<-corHMM(S,gen.data,
                     rate.mat=rate.mat2,
                     node.states="marginal",
                     rate.cat=1,p=cynipgen.ER$rates,
                     root.p=cynipgen.ER$pi)
fit.marginal2$states
#Marginal states are stores in $states. Extract this and ensure the model is consistent
asr2<-fit.marginal2$states
colnames(asr2)<-colnames(rate.mat)
colnames(asr2)<-gsub("/","+",colnames(asr2))
asr2
```

##### Ancestral State Reconstruction

```{r fig.height=10, fig.width=20, fig.align='center'}
#svg("/home/guerbrown/github_local/biol-4386-course-project-guerbrown/output/inqcophy_traitmodeling_CynipGenRecon_guerbrown.svg", width = 20, height = 10)
pp2 <- matrix(0, length(cynipgen), 3, dimnames = list(names(cynipgen), c("Asexual", "Sexual", "Asexual+Sexual")))
plot.phylo(S, type = "fan", cex = 2, align.tip.label = F, lwd = 3.5, label.offset = 0.1, use.edge.length = F, open.angle =  165)
X<-strsplit(setNames(as.character(rate.mat2),names(rate.mat2)),"+", fixed=TRUE)
pies<-matrix(0,Ntip(S),2,
	dimnames=list(S$tip.label,
	c("Asexual","Sexual")))
for(i in 1:Ntip(S)) 
	pies[S$tip.label[i],
		X[[S$tip.label[i]]]]<-
		rep(1/length(X[[S$tip.label[i]]]),
		length(X[[S$tip.label[i]]]))
cols2 <- setNames(c("blue", "red", "yellow"),
	c("Asexual", "Sexual"))
par(fg="black")
pp2 <- matrix(0, length(cynipgen), 3, dimnames = list(names(cynipgen), c("Asexual", "Sexual", "Asexual+Sexual")))
tiplabels(pie=pp2,piecol=cols2,cex=0.15)
par(fg="black")
piecol<-cols2
for(i in 1:ncol(asr2)){
	nn2<-strsplit(colnames(asr2)[i],"+",fixed=TRUE)[[1]]
	if(length(nn)==1) piecol[i]<-cols[nn]
	else if(length(nn2)==2) piecol[i]<-colorRampPalette(cols[nn2])(3)[2]
	else piecol[i]<-"black"
}
names(piecol)<-colnames(asr2)
par(fg="transparent")
nodelabels(pie=asr2,piecol=cols2,cex=0.25)
par(fg="black")
legend("left", c("Asexual", "Sexual", "Asexual+Sexual"), pch = 21, pt.bg = c("blue", "red", "yellow"), pt.cex = 3, title = "Oak Section", cex = 1.5)
#dev.off()
pp2
```

#### Mono/Polythalamous

##### Model Recapitulation

```{r}
mp <- setNames(data$mono.polythalamous, rownames(data))
levels(mp) <- unique(data$monopoly)
mp.ER <- fitpolyMk(S, mp, model = "ER")
xx <- apply(mp.ER$data,1,
            function(x,ss) ss[which(x==1)],
            ss=colnames(mp.ER$data))
xx<-gsub("+","/",xx,fixed=TRUE)
mp.data <- data.frame(Genus_sp = names(xx), habitat = xx)
rate.mat <- mp.ER$index.matrix
rate.mat[rate.mat==0] <- NA
colnames(rate.mat) <- rownames(rate.mat) <- gsub("+","/",
                                             colnames(rate.mat),fixed=TRUE)
ind <- order(colnames(rate.mat))
rate.mat <- rate.mat[ind,ind]
fit.marginal<-corHMM(S,mp.data,
                     rate.mat=rate.mat,
                     node.states="marginal",
                     rate.cat=1,p=mp.ER$rates,
                     root.p=cynipgen.ER$pi)
asr<-fit.marginal$states
colnames(asr)<-colnames(rate.mat)
colnames(asr)<-gsub("/","+",colnames(asr))
```

##### Ancestral State Reconstruction

```{r fig.height=10, fig.width=20, fig.align='center'}
#svg("/home/guerbrown/github_local/biol-4386-course-project-guerbrown/output/inqcophy_traitmodeling_MPRecon_guerbrown.svg", width = 20, height = 10)
plot.phylo(S, type = "fan", cex = 2, align.tip.label = F, lwd = 3.5, label.offset = 0.1, use.edge.length = F, open.angle =  165)
X<-strsplit(setNames(as.character(rate.mat),names(rate.mat)),"+", fixed=TRUE)
pies<-matrix(0,Ntip(S),2,
	dimnames=list(S$tip.label,
	c("Monothalamous","Polythalamous")))
for(i in 1:Ntip(S)) 
	pies[S$tip.label[i],
		X[[S$tip.label[i]]]]<-
		rep(1/length(X[[S$tip.label[i]]]),
		length(X[[S$tip.label[i]]]))
cols2 <- setNames(c("blue", "red", "yellow"),
	c("Monothalamous","Polythalamous"))
par(fg="black")
tiplabels(pie=pp,piecol=cols2,cex=0.15)
par(fg="black")
piecol<-cols2
for(i in 1:ncol(asr)){
	nn<-strsplit(colnames(asr)[i],"+",fixed=TRUE)[[1]]
	if(length(nn)==1) piecol[i]<-cols[nn]
	else if(length(nn)==2) piecol[i]<-colorRampPalette(cols[nn])(3)[2]
	else piecol[i]<-"black"
}
names(piecol)<-colnames(asr)
par(fg="transparent")
nodelabels(pie=asr,piecol=cols2,cex=0.25)
par(fg="black")
legend("bottom", c("Monothalamous","Polythalamous", "Monothalamous+Polythalamous"), pch = 21, pt.bg = c("blue", "red", "yellow"), pt.cex = 3, title = "Oak Section", cex = 1.5)
#dev.off()
```






















