---
title: "Synergus PACo Script"
author: "G. Brown"
date: "2023-02-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### The following code was designed to complete a procrustean analysis of cophylogeny using Synergus inquilines and their hosts in Cynipidae. Synergus trees were made with concatenated data of ultra conserved elements (UCE) and CO1 (made by A. Forbes and M. Zhang). Species were chosen based on conservative estimates (Ward et al., 2023; manuscript in preperation). 

### Set files
##### This should be the only information you need to change in order to recapitulate this analysis.

```{r}
setwd("/home/guerbrown/github_local/biol-4386-course-project-guerbrown/data/PACo")
parasitetree <- "/home/guerbrown/github_local/biol-4386-course-project-guerbrown/data/PACo/inqcophy_PACo_UCECO1SynTree_guerbrown.nex"
hosttree <- "/home/guerbrown/github_local/biol-4386-course-project-guerbrown/data/PACo/inqcophy_PACo_CynipidTree_guerbrown.nex"
assocdata <- "/home/guerbrown/github_local/biol-4386-course-project-guerbrown/data/PACo/inqcophy_PACo_AssociationMatrix_guerbrown.csv"
```

### Dependencies

```{r Library Chunk}
library(paco)
library(ape)
library(vegan)
library(phytools)
library(plotrix)
```

#### Importing the Synergus tree

```{r Synergus Import and Pruning, error = T}
setwd("/home/guerbrown/github_local/biol-4386-course-project-guerbrown/output")
svg("/home/guerbrown/github_local/biol-4386-course-project-guerbrown/output/inqcophy_PACo_SynTree_guerbrown.svg")

#Import
p <- readNexus(parasitetree)

#Fixing tip labels. This is unnecessary but it's nice to make sure everything is plotting correctly

tiplabel_syn <- c("Synergus_sp.6", "S._punctatus_A", "S._campanula_A", "Synergus_sp.3", "Synergus_sp.7", "Synergus_sp.8",
                   "S._walshii_A", "S._erinacei_A", "S._erinacei_B", "S._villosus", "S._magnus", "Synergus_sp.2",
                   "S._ochreus", "S._oneratus_A", "S._oneratus_B", "S._oneratus_C", "S._coniferae", "S._lignicola",
                   "S._laeviventris_A", "S._laeviventris_B", "S._laeviventris_C")
p$tip.label <- tiplabel_syn

#Rooting the phylogeny based on a version of the tree with an outgroup and plotting
root.phylo(p, 1, resolve.root = T)
plot.phylo(p, align.tip.label = T, use.edge.length = F, root.edge = T)
dev.off()
```

#### Importing Cynipidae tree

```{r Cynipid Import and Pruning, error = T}
svg("/home/guerbrown/github_local/biol-4386-course-project-guerbrown/output/inqcophy_PACo_CynipTree_guerbrown.svg", width = 20, height = 10)

#Import
h <- read.nexus(hosttree)

tiplabel_gall <- c("Disholcaspis_quercusglobulus", "Disholcaspis_quercusmamma", "Acraspis_villosa", "Acraspis_quercushirta", "Acraspis_pezomachoides", "Acraspis_erinacei", "Druon_ignotum", "Druon_quercuslanigerum", "Cynips_conspicuus","Philonix_fulvicollis", "Atrusca_quercuscentricola", "Belonocnema_kinseyi", "Andricus_quercusstrobilanus", "Andricus_pisiformis", "Andricus_kingii", "Callirhytis_quercuspunctata", "Callirhytis_quercusventricosa", "Dryocosmus_minusculus", "Amphibolips_quercusostensackenii", "Amphibolips_quercusjuglans", "Amphibolips_cookii")
h$tip.label <- tiplabel_gall

root.phylo(h, 1, resolve.root = T)
plot.phylo(h, align.tip.label = T, use.edge.length = F, root.edge = T)
dev.off()
```

# Importing the association matrix and 

```{r fig.height=10, fig.width=20, fig.align='center', dev='svg'}
#Association matrix import
assoc <- read.table(assocdata, header = T, sep = ",")
#modifying the association matrix to make it readable by the program
hp <- as.data.frame.matrix(assoc)

#Removing underscores from the labels
h$tip.label <- gsub("_", " ", h$tip.label)
p$tip.label <- gsub("_", " ", p$tip.label)
hp$Gallers <- gsub("_", " ", hp$Galler)
hp$Synergus <- gsub("_", " ", hp$Synergus)

#plotting the tangelgram
svg("/home/guerbrown/github_local/biol-4386-course-project-guerbrown/output/inqcophy_PACo_Tanglegram_guerbrown.svg", width = 20, height = 10)
cophyloplot <- cophyloplot(h, p, hp, show.tip.label = T,
lwd = 4, col = make.transparent("black", 0.3), font = 4,
length.line = 0, gap = -1, space = 135, rotate = F,  align.tip.label = T, 
use.edge.length = F, root.edge = T)+
text(5,19.7, "Cynipidae", cex = 1.2)+
text(168, 20.20, "Synergus", cex = 1.2)
dev.off()
```

## PACo

```{r}
#Change the two column association matrix into a binary association matrix. This is necessary to calculate principle coordinates in the statistical section.
hp_binary <- table(hp)
h_dist <- cophenetic(h)
p_dist <- cophenetic(p)

#Prepare PACo data and compute the result with 1,000 permutations.
D <- prepare_paco_data(h_dist, p_dist, hp_binary)
D <- add_pcoord(D)
D <- PACo(D, nperm=1000, seed = NA, method= "r0", proc.warning = F)
```

##### This is the result at 10k permutations
$p
- [1] 0

$ss
- [1] 0.02584893

$n
- [1] 1000

### Residuals

```{r}
residuals_paco(D$proc)
```

### Procrustes plot

```{r}
HP.proc <- procrustes(D$H_PCo, D$P_PCo)
HostX <- HP.proc$X
ParY <- HP.proc$Yrot

#Let's make a few versions of the procrustes plot
#Here's one that has everything on it. If it's overwhelming to study, consider using identify() instead of text(). This will allow you to click and choose which points are labeled. 
svg("/home/guerbrown/github_local/biol-4386-course-project-guerbrown/output/inqcophy_PACo_FullProcPlot_guerbrown.svg", width = 20, height = 10)
plot(HostX, asp=1, pch=120)
points(ParY, pch=1)
arrows(ParY[,1], ParY[,2], HostX[,1], HostX[,2], length=0.10, angle = 90, xpd=T)
HostX <- unique(HP.proc$X)
ParY <- unique(HP.proc$Yrot)
text(ParY[,1], ParY[,2], rownames(ParY), offset=0.3, xpd=F, cex=0.7)
text(HostX[,1], HostX[,2], rownames(HostX),offset=0.3, xpd=F, cex= 0.7) 
dev.off()

#Now, here's one without the lines and just Synergus
svg("/home/guerbrown/github_local/biol-4386-course-project-guerbrown/output/inqcophy_PACo_SynProcPlot_guerbrown.svg", width = 20, height = 10)
plot(HostX, asp=1, pch=120)
points(ParY, pch=1)
HostX <- unique(HP.proc$X)
ParY <- unique(HP.proc$Yrot)
text(ParY[,1], ParY[,2], rownames(ParY), offset=0.3, xpd=F, cex=0.7)
dev.off()

#And here's another with just the Cynipid taxa
svg("/home/guerbrown/github_local/biol-4386-course-project-guerbrown/output/inqcophy_PACo_CynProcPlot_guerbrown.svg", width = 20, height = 10)
plot(HostX, asp=1, pch=120)
points(ParY, pch=1)
HostX <- unique(HP.proc$X)
ParY <- unique(HP.proc$Yrot)
text(HostX[,1], HostX[,2], rownames(HostX),offset=0.3, xpd=F, cex= 0.7)
dev.off()
```

### Barplot of links.

```{r fig.height=7.5, fig.width=20, fig.align='center', dev='svg'}
NLinks = 25
SQres <- sort(residuals_paco(D$proc))
SQres.jackn <- SQres**2
SQres <- (residuals (HP.proc)**2)
SQres.jackn <- SQres.jackn*(-(NLinks-1))
SQres.jackn <- SQres.jackn*-1
SQres <- SQres*NLinks
SQres <- as.matrix(SQres)
SQres.jackn <- as.data.frame(SQres.jackn)
SQres.jackn <- t(apply(SQres.jackn, 1, "+", SQres))
phi.mean <- apply(SQres.jackn, 2, mean, na.rm = TRUE)
phi.UCI <- apply(SQres.jackn, 2, sd, na.rm = TRUE)
t.critical = qt(0.975, NLinks-1) 
phi.UCI <- phi.mean + t.critical * phi.UCI/sqrt(NLinks)
phi.UCI

svg("/home/guerbrown/github_local/biol-4386-course-project-guerbrown/output/inqcophy_PACo_ResBarplot_guerbrown.svg", width = 20, height = 7.5)
par(mar = c(7, 5, 3, 6))
pat.bar <- barplot(sort(phi.mean), names.arg = "", space = 0.2, col="white", ylab= "Squared residuals", ylim=c(0, max(phi.UCI)),cex.lab=1)
segments(pat.bar, sort(phi.mean), pat.bar, sort(phi.UCI), lwd = 1.5)
arrows(pat.bar, sort(phi.mean), pat.bar, sort(phi.UCI), lwd = 1.5, angle = 90, code = 3, length = 0.05)
abline(a=median(phi.mean), b=0, lty=2) 
text(pat.bar, par("usr")[3] - 0.001, srt = 330, adj = 0, labels = rownames(SQres.jackn), xpd = TRUE, font = 3, cex=0.65, las = 2)
text(22.32, 0.05, '*', cex = 3)
text(25.93, 0.053, '*', cex = 3)
dev.off()
```

### Significance of squared residuals using parafit link
### Lingoes: In the Lingoes (1971) procedure, a constant c1, equal to twice absolute value of the largest negative value of the original principal coordinate analysis, is added to each original squared distance in the distance matrix, except the diagonal values. A new principal coordinate analysis, performed on the modified distances, has at most (n-2) positive eigenvalues, at least 2 null eigenvalues, and no negative eigenvalue.

### Cailliez: In the Cailliez (1983) procedure, a constant c2 is added to the original distances in the distance matrix, except the diagonal values. The calculation of c2 is described in Legendre and Legendre (1998). A new principal coordinate analysis, performed on the modified distances, has at most (n-2) positive eigenvalues, at least 2 null eigenvalues, and no negative eigenvalue.

## Note to Andrew: Calliez and Lingoes corrections add a constant to each value that negates negative eigenvalues. This is to make sure all values can be plotted in real space but it can inflate the final statistics. So, Hutchinson et al.,2017 recommended taking the square root of the eigenvalues to convert everything to positive. I've included both here. We can discuss which one is a better fit for our analysis later.

### Sqrt of eigenvalues manual correction

```{r}
h_matrix <- compute.mr(h)
h_df <- as.data.frame(h_matrix)
host.D <- cophenetic(h)
para.D <- cophenetic(p)
parafit <- parafit(sqrt(host.D), sqrt(para.D), hp_binary, test.links = T, correction = "none", nperm = 1000)
print(parafit)
```

### Cailliez Correction

```{r}
parafit <- parafit(host.D, para.D, hp_binary, test.links = T, correction = "cailliez", nperm = 1000)
print(parafit)
```

### Lingoes correction

```{r}
parafit <- parafit(host.D, para.D, hp_binary, test.links = T, correction = "lingoes", nperm = 1000)
print(parafit)
```








